Gaps in stellar streams as a result of globular cluster fly-bys

### 2025-05-15 paper/main.tex
uncommented plots, added label, fixed ??

  \date{SUBMITTED TO A\&A 27 JAN 2025; Accepted XXX}

@@

\subsection{Mass loss}

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{decomposition-monte-carlo-009-with-3-gaps-domidpoint-shift.png}
      \caption{Density maps of Palomar~5's stream in the tail coordinate system. The color scale represents normalized particle counts (total: 100,000). The top panel shows the \texttt{full} simulation with three gaps on the stream's right-hand side. The next three panels depict simulations with identical initial conditions but exclude the gravitational influence of all clusters except those forming a given gap. The \texttt{Reference} simulation omits the influence from other globular clusters. The bottom panel compares Palomar~5's observed stream length to the \texttt{Reference} simulation, using the same Monte Carlo realization as Fig.~\ref{fig:stream_on_sky} and \texttt{Sampling~009} of Fig.~\ref{fig:gallery2}.}
      \label{fig:decomposition}
    \end{figure*}

@@

  The impact of globular clusters' close interactions with stellar streams has also been the object of another recent paper by \citet{2022ApJ...941..129D}, who concluded that the chance for GD-1 gaps, which have been reported in several works \citep[see, for example, ][]{2019ApJ...880...38B,2018MNRAS.477.1893D,2020AAS...23533607D} to be produced by globular clusters is very low. This result is not necessarily in contradiction with ours since GD-1 has a pericenter which is almost twice that of Palomar~5 \citep[see, for example][]{2019MNRAS.486.2995M}. As discussed in Sect.~\ref{sect:geometry}, the number, $N$, of close encounters that lead to gap creation is anti-correlated with the distance $r$ to the Galactic center. If we naively use the same radial dependence of gaps from Palomar~5 for GD-1 by swapping a pericenter from 6~kpc to 16~kpc, we would reduce the number of gap-favorable impacts by more than a factor of 2. Moreover, we note that \citet{2022ApJ...941..129D} pre-selects the globular clusters that could have experienced a close encounter with the GD-1 stream by selecting only clusters that pass at a distance less than 0.5 kpc from the stream, having a relative velocity smaller than  300 km/s. As shown in our Fig.~\ref{fig:impact_geometry_statistics}, bottom-left panel, in the case of Palomar~5, this choice would lead to excluding most of the encounters favorable to gap creation, which turn out to have relative velocities above 300 km/s. It would be interesting to repeat a similar study as the one made by \citet{2022ApJ...941..129D} for GD-1, imposing no selection on possible candidate clusters.\\

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-000.png}
      % \includegraphics[width=\linewidth]{tau-profile-monte-carlo-000.png}
      % \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-001.png}
      % \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-002.png}
      % \includegraphics[width=\linewidth]{tau-profile-monte-carlo-002.png}
      \caption{First gap gallery. See the text in Appendix~\ref{sec:gallery_of_gaps} for a description.}
      \label{fig:gallery0}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-003.png}      
      % \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-004.png}
      % \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-005.png}
      % \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-006.png}
      % \includegraphics[width=\linewidth]{tau-profile-monte-carlo-006.png}
      \caption{Second gap gallery}
      \label{fig:gallery1}
      \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-007.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-008.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-009.png}      
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-010.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-011.png}
    %   \caption{Third gap gallery}
    %   \label{fig:gallery2}
    %   \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-012.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-013.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-014.png}      
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-015.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-016.png}
    %   \caption{Fourth gap gallery}
    %   \label{fig:gallery3}
    %   \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-017.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-018.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-019.png}      
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-020.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-021.png}
    %   \caption{Fifth gap gallery}
    %   \label{fig:gallery4}
    %   \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-022.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-023.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-024.png}   
    %   \includegraphics[width=\linewidth]{tau-profile-monte-carlo-024.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-025.png}
    %   \caption{Sixth gap gallery}
    %   \label{fig:gallery5}
    %   \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-026.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-027.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-028.png}
    %   \includegraphics[width=\linewidth]{tau-profile-monte-carlo-028.png}
    %   \caption{Seventh gap gallery}
    %   \label{fig:gallery6}
    %   \end{figure*}

    % \begin{figure*}
    %   \centering      
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-029.png}    
    %   \includegraphics[width=\linewidth]{tau-profile-monte-carlo-029.png}  
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-030.png}
    %   \includegraphics[width=\linewidth]{tau-profile-monte-carlo-030.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-031.png}
    %   \caption{Eighth gap gallery}
    %   \label{fig:gallery7}
    %   \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-032.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-033.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-034.png}      
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-035.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-036.png}      
    %   \caption{Ninth gap gallery}
    %   \label{fig:gallery8}
    %   \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-037.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-038.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-039.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-040.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-041.png}      
    %   \caption{Tenth gap gallery}
    %   \label{fig:gallery9}
    % \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-042.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-043.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-044.png}
    %   \includegraphics[width=\linewidth]{tau-profile-monte-carlo-044.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-045.png}
    %   \caption{Eleventh gap gallery}
    %   \label{fig:gallery010}
    % \end{figure*}

    % \begin{figure*}
    %   \centering
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-046.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-047.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-048.png}
    %   \includegraphics[width=\linewidth]{tau-profile-monte-carlo-048.png}
    %   \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-049.png}
    %   \caption{Twelfth gap gallery}
    %   \label{fig:gallery11}
    % \end{figure*}

>>>>

  \date{SUBMITTED TO A\&A XXX; Accepted XXX}

@@

\subsection{Mass loss}\label{sec:mass_loss}

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{decomposition-monte-carlo-009-with-3-gaps-domidpoint-shift.png}
      \caption{Density maps of Palomar~5's stream in the tail coordinate system, where $x\prime$ is the integrate arclength along the cluster's orbit and $y\prime$ is the distance within the orbital plane. The color scale represents normalized particle counts (total: 100,000). The top panel shows the \texttt{full} simulation with three gaps on the stream's right-hand side. The next three panels depict simulations with identical initial conditions but exclude the gravitational influence of all clusters except those forming a given gap. The \texttt{Reference} simulation omits the influence from other globular clusters. The bottom panel compares Palomar~5's observed stream length to the \texttt{Reference} simulation, using the same Monte Carlo realization as Fig.~\ref{fig:stream_on_sky} and \texttt{Sampling~009} of Fig.~\ref{fig:gallery2}.}
      \label{fig:decomposition}
    \end{figure*}

@@

  The impact of globular clusters' close interactions with stellar streams has also been the object of another recent paper by \citet{2022ApJ...941..129D}, who concluded that the chance for GD-1 gaps, which have been reported in several works \citep[see, for example, ][]{2019ApJ...880...38B,2018MNRAS.477.1893D,2020AAS...23533607D} to be produced by globular clusters is very low. This result is not necessarily in contradiction with ours since GD-1 has a pericenter which is almost twice that of Palomar~5 \citep[see, for example][]{2019MNRAS.486.2995M}. As discussed in Sect.~\ref{sect:geometry}, the number, $N$, of close encounters that lead to gap creation is anti-correlated with the distance $r$ to the Galactic center. If we naively use the same radial dependence of gaps from Palomar~5 for GD-1 by swapping a pericenter from 6~kpc to 16~kpc, we would reduce the number of gap-favorable impacts by more than a factor of 2. Moreover, we note that \citet{2022ApJ...941..129D} pre-selects the globular clusters that could have experienced a close encounter with the GD-1 stream by selecting only clusters that pass at a distance less than 0.5 kpc from the stream, having a relative velocity smaller than  300 km/s. As shown in our Fig.~\ref{fig:impact_geometry_statistics}, bottom-left panel, in the case of Palomar~5, this choice would lead to excluding most of the encounters favorable to gap creation, which turn out to have relative velocities above 300 km/s, \textbf{reducing the total to just 19 gaps}. It would be interesting to repeat a similar study as the one made by \citet{2022ApJ...941..129D} for GD-1, imposing no selection on possible candidate clusters.\\

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-000.png}
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-000.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-001.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-002.png}
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-002.png}
      \caption{First gap gallery. See the text in Appendix~\ref{sec:gallery_of_gaps} for a description.}
      \label{fig:gallery0}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-003.png}      
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-004.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-005.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-006.png}
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-006.png}
      \caption{Second gap gallery}
      \label{fig:gallery1}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-007.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-008.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-009.png}      
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-010.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-011.png}
      \caption{Third gap gallery}
      \label{fig:gallery2}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-012.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-013.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-014.png}      
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-015.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-016.png}
      \caption{Fourth gap gallery}
      \label{fig:gallery3}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-017.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-018.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-019.png}      
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-020.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-021.png}
      \caption{Fifth gap gallery}
      \label{fig:gallery4}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-022.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-023.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-024.png}   
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-024.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-025.png}
      \caption{Sixth gap gallery}
      \label{fig:gallery5}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-026.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-027.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-028.png}
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-028.png}
      \caption{Seventh gap gallery}
      \label{fig:gallery6}
      \end{figure*}

    \begin{figure*}
      \centering      
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-029.png}    
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-029.png}  
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-030.png}
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-030.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-031.png}
      \caption{Eighth gap gallery}
      \label{fig:gallery7}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-032.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-033.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-034.png}      
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-035.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-036.png}      
      \caption{Ninth gap gallery}
      \label{fig:gallery8}
      \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-037.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-038.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-039.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-040.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-041.png}      
      \caption{Tenth gap gallery}
      \label{fig:gallery9}
    \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-042.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-043.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-044.png}
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-044.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-045.png}
      \caption{Eleventh gap gallery}
      \label{fig:gallery010}
    \end{figure*}

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-046.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-047.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-048.png}
      \includegraphics[width=\linewidth]{tau-profile-monte-carlo-048.png}
      \includegraphics[width=\linewidth]{gallery_of_gaps_monte-carlo-049.png}
      \caption{Twelfth gap gallery}
      \label{fig:gallery11}
    \end{figure*}
